Comparison of Spatial Partitioning Strategies

### 2025-02-28 paper/paper.tex
paper: Finished the first draft.

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage{svg}
\usepackage{float}

@@

\begin{titlepage}
    \centering
    \vspace*{1cm}

@@

    \vspace{1cm}
    {\today}
\end{titlepage}

@@

\section{Purpose \& Scope}
The purpose of this paper is to find the most optimal strategy for dividing up
space to improve the speed of spatial querying of nearby objects.

This study focuses mainly on CPU performance and runtime in exchange for less
memory efficient implementation. Therefore time based evaluations of insertion,
query time, clearing time and general collision testing time will be performed.

The strategies are going to be evaluated on time based metrics - insertion,
query, clearing and time spend doing collision testing. These evaluations are
going to take place with two different distribution strategies of the objects
though the same seed will be used for all tests so the pseudo-random numbers
stay the same.

@@

\section{Discussion}

@@

>>>>

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage{svg}
\usepackage{float}
\usepackage{hyperref}

@@

\pagenumbering{alph}
\begin{titlepage}
    \centering
    \vspace*{1cm}

@@

    \vspace{1cm}
    {\today}
\end{titlepage}
\pagenumbering{arabic}

@@

\section{Purpose \& Scope}
\label{sec:purpose-and-scope}
The purpose of this paper is to find the most optimal strategy for dividing up
space to improve the speed of spatial querying of nearby objects.

This study focuses mainly on CPU performance and runtime in exchange for less
memory efficient implementation. Therefore time based evaluations of insertion,
query time and general collision testing time will be performed.

The strategies are going to be evaluated on time based metrics - insertion,
query and time spend doing collision testing. These evaluations are going to
take place with two different distribution strategies of the objects though the
same seed will be used for all tests so the pseudo-random numbers stay the same.

@@

\section{Discussion}
The design choices for this project was entirely focused on CPU performance
leaving memory as an afterthought. This was done in order to mimic a modern game
running on consumer hardware where memory is often an abundant resource.
Everything inside the spatial partitioning strategies were allocated when the
structure was initialized so no heap allocations were being issued in the middle
of a hot loop. There are, however, two notable exception to this philosophy.
When running queries I used vectors, which have to allocate memory on the heap,
when doing queries. Looking back, there could be more efficient ways to handle
this. Instead of having the query function handle the memory management, let the
called handle that complexity. Have the caller allocate an array with sufficient
enough size and pass that as a parameter for the function to modify. Though this
would work it's not really a good solution, in my subjective opinion. Because
there is guarantee of how many objects could be returned. The better solution
would be to use a bump, arena or block allocator to further speed up this
operation instead of using the general purpose malloc and free pattern provided
by the C standard library. The other notable exception is the benchmarking tool
I made for this. It uses a hash map to store how many times a function was run,
the total time it ran and the average runtime per call. Because of my underlying
implementation of the hash map it will require memory allocation when enough
entries has been added. Not only that, but it was a nested structure which
created a new hash map for each benchmark scope.

The benchmarks are a thing I would have done differently if I would be redoing
this project. There are a few decisions that I can't justify now. One of these
weird choices is how I chose to benchmark insertions. Instead of running the
benchmark for each object inserted I measured to total time taken to insert all
objects. This doesn't sound weird until you look at how the rest of the
benchmarks are measured. The queries are measured once per object instead of
encapsulating all objects in a single benchmark. The insert metric is the only
one which differs in this regard. It would be an easy fix code wise, but having
to rewrite the paper based off of a new metric would be too much work to justify
the change.

I found that although quadtrees outperform all the other strategies in this
particular program, it may not be the most suitable for all programs. Both grids
and quadtrees limit you to a fixed known world size which could be a problem for
larger games or simulations. If the quadtree is covering a massive region of the
world then that would introduce significant overhead because there would have to
be a lot of subdivisions happening. Chunking could fix this issue though there
are other strategies more suitable for this. The spatial hashing grid solves
this issue at the cost of performance. Grids come with the same world size
restriction that the quadtree does but has the benefit of being extremely
trivial to implement. At with almost all questions the answer to 'In what use
case are the different algorithms most useful?' the answer is: it depends. For
static objects which cover a known area, the quadtree is by far the most useful.
For dynamic objects which move a lot the spatial hashing grid is more suitable.
And lastly, the simple grid is really handy when you quickly want to increase
the speed of a system without having to spend a lot of time writing the
optimization strategy, which works really well for applications with a smaller
amount of objects. A combination of multiple strategies could be used to gain
the benefit of them, like utilizing the quadtree for static terrain while having
all dynamic objects be stored in a spatial hashing gird.

Then at what point is it worth implementing a more advanced data structure? Well
that also depends heavily on your particular use case. If the problem only
involves checking a handful of objects then the naive solution works just fine.
As seen in \ref{fig:naive} the naive solution is rarely viable as the object
count grows. The only real way to know when it is worth to implement a new
optimization strategy is to measure your current solution and make a decision
based off of that.

@@

\section{Appendix}
The source code written to get these metrics, alongside the python scripts used
to generate the graphs, can be found at
\url{https://github.com/faintforge/spatial-partitioning}.
